---
title: "Vignette for AgslPredict"
author: "Geoffrey R. Hosack and Maud El-Hachem"
date: "2025-04-08"
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Vignette for AgslPredict}
-->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>"
)
```

# Outline

This vignette documents the code and data provided by package **AgslPredict** as supplementary material for the paper ``Prediction of mosquito vector abundance for three species in the Anopheles gambiae complex'', <https://doi.org/10.1101/2025.03.04.641552>. The underlying response data requires multiple downloads and some of the analyses were implemented on advanced computing facilities. The scripts and previously downloaded data are therefore stored in top level directories of this package, which are referenced in this vignette. The scripts refer to files in the different directories to build dataframes and execute analysis that may use high performance computing facilities.

* [01 Relative Abundance Data] 
* [02 Relative Abundance Analysis] 
* [03 Abundance Data] 
* [04 Abundance Analysis]
* [05 Prediction]

Please see file LICENSE included with the **AgslPredict** package for license information, data sources and citations.

# 01 Relative Abundance Data

The raw data used for relative abundance are available at <https://vectorbase.org/common/downloads/popbio-map-legacy/VectorBase-popbio-map-rel64-July2023-Samples.csv.gz>. The R script `RelativeAbundanceDataPrep.R` that downloads and processes these data for analysis is available in this package by executing
```{r ra, eval = FALSE}
system.file(package = "AgslPredict", "01-RelativeAbundanceData/")
```

VectorBase identified species for many samples of *An. gambiae* sensu lato (s.l.). Samples identified by chromosomal forms were related to species: "Anopheles gambiae chromosomal form Bamako" and "Anopheles gambiae chromosomal form Savanna" was assigned to "Anopheles gambiae sensu stricto", and "Anopheles gambiae chromosomal form Mopti" was assigned to "Anopheles coluzzii". The three species retained for analysis were "Anopheles arabiensis", "Anopheles coluzzii" and "Anopheles gambiae sensu stricto".

Samples outside the extent bounded by latitudes between 34S and 20N and longitudes between 20W and 52E were filtered out. Other samples initially filtered out of the database were those with no information for date of collection (1\%) or only years (1\%). Of the remainder, about 14\% were reported to the month, 52\% were reported to the day, and 23\% were reported as a duration between two dates. Of the latter, 94\% of the durations between dates were one day or less (i.e., collection during a day or overnight). The start and end date of these records were extracted. Samples reported to the level of one month were assigned start and end dates that corresponded with the start and end date of the reported month. Remaining samples with durations greater than 31 days (10\%) were filtered out of the dataset. Samples with initial dates occurring before 2001 were omitted.

Samples were then identified by project ID, latitude and longitude, collection date and collection protocol to define the unit of analysis. The frequency counts of the three species identified for each unit sample were then recorded. Sample units with shared collection dates and collection methods were then aggregated to 5 km x 5 km grid cells across the spatial extent of the analysis. The grid cell file is found at 
```{r gc,eval=FALSE}
system.file(package = "AgslPredict", "covariates_spatial/activeAfrica.tif")
```

## Covariate Data for Relative Abundance

The relative abundance data incorporates meteorological information with download script located at

```{r mcvar,eval=FALSE}
system.file(package = "AgslPredict", 
            "meteo_for_relative_abundance/meteo_download_RelativeAbundance.R")
```

Spatial covariate data are stored at

```{r dat1, eval = FALSE}
system.file(package = "AgslPredict", "covariates_spatial")
```

Vector intervention data should be downloaded and processed by scripts `ITN_process.R` and `IRS_process.R` found at
```{r, vi, eval = FALSE}
system.file(package = "AgslPredict", "vector_intervention")
```

The covariate data for the relative abundance data analysis are integrated into the dataset using the script found at
```{r rac,eval=FALSE}
system.file(package = "AgslPredict", "01-RelativeAbundanceData/RelativeAbundanceDataCovariates.R")
```

# 02 Relative Abundance Analysis

The R script that analyses the relative abundance data for analysis is located at
```{r raa, eval = FALSE}
system.file(package = "AgslPredict", 
            "02-RelativeAbundanceAnalysis/relative_abundance_analysis_1m.R")
```

This script reads in the relative abundance data and covariates. A series of multinomial logit GLMs are then fit using a backward stepwise BIC procedure on two groups of models 1 month windowed meteorological covariates without transformations.

The best fitting model included all possible covariates and most interactions and squared terms. The directory includes spatial plots of residuals by species, model predictions by species assuming most recent survey year and observed data by species. 

The best fitting model object is stored at
```{r rab, eval = FALSE}
system.file(package = "AgslPredict", 
            "02-RelativeAbundanceAnalysis/multinomialGLMfit1m.rds")
```


# 03 Abundance Data

The R script that analyses the abundance data and covariates for analysis is located at
```{r adat, eval = FALSE}
system.file(package = "AgslPredict", 
            "03-AbundanceData/AbundancePrep.R")
```

This script creates a file ("03-AbundanceData/abundance.rds") that is sourced by the compilation script for meteorological covariates at

```{r apdat, eval = FALSE}
system.file(package = "AgslPredict", 
            "meteo_for_abundance/meteo_abundance_download.R")
```

Human population data should be downloaded and processed by scripts `demographic_process.R` found at

```{r pop, eval = FALSE}
system.file(package = "AgslPredict", 
            "human_pop/demographic_process.R")
```

The abundance data, meteorological covariate data, human population covariates and spatial covariates are integrated into a single dataset by the script located at
```{r spac, eval = FALSE}
system.file(package = "AgslPredict", 
            "03-AbundanceData/AbundanceCovarPrep.R")
```

# 04 Abundance Analysis

The abundance analysis uses the script

```{r aanalysis, eval = FALSE}
system.file(package = "AgslPredict", 
            "04-AbundanceAnalysis/agsl_negbinom.R")
```

The plot of predicted abundance with varying human population and precipitation uses the script
```{r apredict, eval = FALSE}
system.file(package = "AgslPredict",
            "04-AbundanceAnalysis/predict_abundance_plot.R")
```

# 05 Prediction

## Meteorological Data for Prediction

The meteorological data files used for prediction were sourced using the script at
```{r dat3s, eval = FALSE}
system.file(package = "AgslPredict",
            "meteo_prediction_raw/download_from_power.R")
```

## Spatio-Temporal Predictions

The predictions were processed on high performance computing by the following scripts. The daily predictions across sub-Saharan Africa for years from 2002 to 2020 benefit from parallel processing to reduce user runtime. The script batches simulations by cells using the variable `id`, which if set to `NULL` inherits command line arguments available when the `R` session is invoked.

```{r pred1, eval = FALSE}
system.file(package = "AgslPredict",
            "05-Prediction/prediction_tot_relabund.R")
```

The above script sources the prediction function at 
```{r pred2, eval = FALSE}
system.file(package = "AgslPredict",
            "05-Prediction/pXPrediction_function.R")
```

The following script defines the temperature threshold mask for predictions at
```{r pred2a, eval = FALSE}
system.file(package = "AgslPredict",
            "05-Prediction/tempMask.R")
```

The script for plots of predictions is located at
```{r pred3, eval = FALSE}
system.file(package = "AgslPredict",
            "05-Prediction/prediction_plots.R")
```

Niche overlap and competition estimates are located at
```{r pred4, eval = FALSE}
system.file(package = "AgslPredict",
            "05-Prediction/analysisSpec.R")
```

## Covariate Plots

Visualisations for the meteorological data computed by the preprocessing script on high performance located at
```{r metprep,eval=FALSE}
system.file(package = "AgslPredict",
            "covar_plots/calc_metavg.R")
```

Summary plots of the meteorological data are created by the script
```{r metplots,eval=FALSE}
system.file(package = "AgslPredict", "covar_plots/met_plots.R")
```

Plots for spatial covariates are created by the script located at
```{r spatplots,eval=FALSE}
system.file(package = "AgslPredict", "covar_plots/spatial_plots.R")
```

The human population data were downloaded from <https://landscan.ornl.gov/> and processed by scripts located at

```{r popdat, eval = FALSE}
system.file(package = "AgslPredict", 
            "human_pop/demographic_process.R")
```

Plots for the human population data are created by the scripts located at
```{r popplots, eval= FALSE}
system.file(package = "AgslPredict", "covar_plots/pop_plots.R")
```

The vector intervention data were downloaded from <https://malariaatlas.org/> and processed by scripts located at
```{r vecdat, eval=FALSE}
system.file(package = "AgslPredict", "vector_intervention")
```

Plots for the vector intervention data are created by the scripts located at
```{r vecplots, eval= FALSE}
system.file(package = "AgslPredict", "covar_plots/intervention_plots.R")
```
